Characterization and evolution of hourly extreme precipitation in France using an Explicit-Convection Regional Climate Model

### 2025-10-15 article/article.qmd
Articles v1

AROME indicates positive trends (+25% to >+150%) in the Rhône Valley and the Prealps in February (**FEB**), over the western Mediterranean arc (Roussillon–Lower Languedoc) in March (**MAR**), and the eastern arc (Azur Basin) in November (**NOV**). Station data confirm this pattern, markedly strengthening the signal in the Rhône Valley in February, with trends approximately twice as high as those estimated by AROME. In June (**JUN**), both AROME and station data show a predominantly increasing signal across the entire territory, with local trends sometimes exceeding +150%. However, the extremely low spatial correlation between the two datasets (*r* = 0.07) indicates that, while the average sign converges, the localization of trends is inconsistent/noisy. This discrepancy, combined with high summer convective variability and the short length of monthly series (1990–2022), calls for caution.

@@

    % --- Ligne 2 ---
    \textbf{\small JUI} & \textbf{\small NOV} \\[-0.5mm]
    \small $r =$ {`{python} f'{df_trend_pluie_jui['r'].iloc[0].item():.2f}'`} (n={`{python} f'{df_trend_pluie_jui['n'].iloc[0].item()}'`}) &
    \small $r =$ {`{python} f'{df_trend_pluie_nov['r'].iloc[0].item():.2f}'`} (n={`{python} f'{df_trend_pluie_nov['n'].iloc[0].item()}'`}) \\[-0.5mm]
    \small $ME =$ {`{python} f'{df_trend_pluie_jui['me'].iloc[0].item():+.2f}'`}\% &
    \small $ME =$ {`{python} f'{df_trend_pluie_nov['me'].iloc[0].item():+.2f}'`}\% \\[-0.5mm]
    \includegraphics[width=\linewidth]{figures/trend_horaire_pluie_jui.pdf} &
    \includegraphics[width=\linewidth]{figures/trend_horaire_pluie_nov.pdf} \\

@@

To objectify these divergences and assess the robustness of the signals, we now examine the spatial correlations between AROME and station data, by season and month, in order to identify where agreement is structural and where it is mostly noise. At the daily scale (1959–2022), trend correlations range from 0.14 (**HYDRO**) to 0.32 (**OND**) for seasons and from 0.07 (**JUN**) to 0.51 (**DEC**) for months, with several months reaching or exceeding 0.40 (**JAN**, **MAR**, **AUG**, **NOV**) (Figure 8). Over the shorter period (1990–2022), seasonal values span 0.19–0.40 and monthly values 0.02–0.47 (**JAN** 0.47; **DEC** 0.35). At the hourly scale (1990–2022), seasons exhibit weak correlations ranging from 0.00 (**JAS**) to 0.09 (**OND**), and monthly correlations extend from 0.02 (**SEP**, **OCT**) to 0.42 (**FEB**).

@@

**Hourly data.** The signals are much less robust at the hourly scale than at the daily scale. They show very heterogeneous trends, often weakly or not significant, and sometimes spatially inconsistent between AROME and stations (in June, *r* = 0.07). This aligns with the @IPCC2021 finding of low confidence in a global increase in sub-daily extremes, due to the lack of long, dense, and homogeneous series. The marked increases in February (Rhône Valley, Pre-Alps), March (western Mediterranean arc), and November (eastern Mediterranean arc) are compatible with dynamic and thermo-marine regimes conducive to Mediterranean/transitional episodes (end of winter, autumn interseason). The signal doubled by stations in February in the Rhône Valley highlights AROME’s smoothing and amplitude bias for hourly extremes. However, the restricted period (1990–2022) offers little statistical power, and interannual convective variability can mask or mimic a trend. In June, the widespread increase (up to +150%), but without spatial agreement between AROME and stations, shows local convective variability (very localized storms) and errors of point vs. grid representativeness become dominant. This is an example of the limits of monthly inference for convective extremes. It should be noted that this does not result from an isolated anomaly or a single extreme year. A sensitivity analysis showed that contrasts persist when the year with the maximum precipitation is excluded. They therefore reflect a structural signal linked to the succession of several marked extreme events over the last decade, reinforcing recent trends rather than a punctual temporal artifact.

>>>>

AROME indicates positive trends (+25% to >+150%) in the Rhône Valley and the Prealps in February (**FEB**), over the western Mediterranean arc (Roussillon–Lower Languedoc) in March (**MAR**), and the eastern arc (Azur Basin) in November (**NOV**). Station data confirm this pattern, markedly strengthening the signal in the Rhône Valley in February, with trends approximately twice as high as those estimated by AROME. In June (**JUN**), both AROME and station data show a predominantly increasing signal across the entire territory, with local trends sometimes exceeding +150%. However, the extremely low spatial correlation between the two datasets (*r* = 0.10) indicates that, while the average sign converges, the localization of trends is inconsistent/noisy. This discrepancy, combined with high summer convective variability and the short length of monthly series (1990–2022), calls for caution.

@@

    % --- Ligne 2 ---
    \textbf{\small JUN} & \textbf{\small NOV} \\[-0.5mm]
    \small $r =$ {`{python} f'{df_trend_pluie_jui['r'].iloc[0].item():.2f}'`} (n={`{python} f'{df_trend_pluie_jui['n'].iloc[0].item()}'`}) &
    \small $r =$ {`{python} f'{df_trend_pluie_nov['r'].iloc[0].item():.2f}'`} (n={`{python} f'{df_trend_pluie_nov['n'].iloc[0].item()}'`}) \\[-0.5mm]
    \small $ME =$ {`{python} f'{df_trend_pluie_jui['me'].iloc[0].item():+.2f}'`}\% &
    \small $ME =$ {`{python} f'{df_trend_pluie_nov['me'].iloc[0].item():+.2f}'`}\% \\[-0.5mm]
    \includegraphics[width=\linewidth]{figures/trend_horaire_pluie_jui.pdf} &
    \includegraphics[width=\linewidth]{figures/trend_horaire_pluie_nov.pdf} \\

@@

To objectify these divergences and assess the robustness of the signals, we now examine the spatial correlations between AROME and station data, by season and month, in order to identify where agreement is structural and where it is mostly noise. At the daily scale (1959–2022), trend correlations range from 0.14 (**HYDRO**) to 0.32 (**OND**) for seasons and from 0.10 (**JUN**) to 0.51 (**DEC**) for months, with several months reaching or exceeding 0.40 (**JAN**, **MAR**, **AUG**, **NOV**) (Figure 8). Over the shorter period (1990–2022), seasonal values span 0.19–0.40 and monthly values 0.02–0.47 (**JAN** 0.47; **DEC** 0.35). At the hourly scale (1990–2022), seasons exhibit weak correlations ranging from 0.00 (**JAS**) to 0.09 (**OND**), and monthly correlations extend from 0.02 (**SEP**, **OCT**) to 0.42 (**FEB**).

@@

**Hourly data.** The signals are much less robust at the hourly scale than at the daily scale. They show very heterogeneous trends, often weakly or not significant, and sometimes spatially inconsistent between AROME and stations (in June, *r* = 0.10). This aligns with the @IPCC2021 finding of low confidence in a global increase in sub-daily extremes, due to the lack of long, dense, and homogeneous series. The marked increases in February (Rhône Valley, Pre-Alps), March (western Mediterranean arc), and November (eastern Mediterranean arc) are compatible with dynamic and thermo-marine regimes conducive to Mediterranean/transitional episodes (end of winter, autumn interseason). The signal doubled by stations in February in the Rhône Valley highlights AROME’s smoothing and amplitude bias for hourly extremes. However, the restricted period (1990–2022) offers little statistical power, and interannual convective variability can mask or mimic a trend. In June, the widespread increase (up to +150%), but without spatial agreement between AROME and stations, shows local convective variability (very localized storms) and errors of point vs. grid representativeness become dominant. This is an example of the limits of monthly inference for convective extremes. It should be noted that this does not result from an isolated anomaly or a single extreme year. A sensitivity analysis showed that contrasts persist when the year with the maximum precipitation is excluded. They therefore reflect a structural signal linked to the succession of several marked extreme events over the last decade, reinforcing recent trends rather than a punctual temporal artifact.
